A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2024-01-28 content/manuscript.tex
Updates from Overleaf

\author[%
1,2%
]{Roman Ludwig}
\author[%
1,2%
]{Julian Brönnimann}
\author[%
1,2%
]{Yoel Perez Haas}
\author[%
1,2%
]{Esmée Lauren Looman}
\author[%
2%
]{Panagiotis Balermpas}
\author[%
11%
]{Sergi Benavente}
\author[%
3,4,7%
]{Adrian Schubert}
\author[%
8%
]{Dorothea Barbatei}
\author[%
8%
]{Laurence Bauwens}
\author[%
2%
]{Jean-Marc Hoffmann}
\author[%
3%
]{Olgun Elicin}
\author[%
6,10%
]{Matthias Dettmer}
\author[%
2%
]{Bertrand Poumayou}
\author[%
4,5%
]{Roland Giger}
\author[%
8%
]{Vincent Grégoire}
\author[%
1,2%
]{Jan Unkelbach}

@@

Head and neck squamous cell carcinomas (HNSCC) frequently spread through
the lymphatic system
\autocite{lindberg_distribution_1972,woolgar_histological_1999}. Current
diagnostic imaging modalities are unable to detect microscopic lymph
node metastases \autocite{snyder_petct_2021,strohl_petct_2021}. To avoid
nodal recurrences, large volumes in the neck, which are at risk of
harbouring occult disease, are irradiated electively. Guidelines about
which lymph node levels (LNLs) to irradiate
\autocite{biau_selection_2019} are currently not based on a patient's
individual risk, but on the overall prevalence of nodal disease as
reported in the literature
\autocite{lindberg_distribution_1972,woolgar_histological_1999}.

To personalize the prediction of the risk for occult disease, given a
patient's individual diagnosis, we published

@@

Such a personalized risk prediction may allow clinicians to safely
reduce the elective clinical target volume (CTV-N) and thus reduce
side-effects that degrade the patient's quality of life
\autocite{batth_practical_2014}.

Here, we extend the previous work by incorporating the primary tumor
location (specified as ICD-10 code) into the model of lymphatic tumor
progression, focusing on tumors in the oropharynx and the oral cavity.
HNSCC patients with primary tumors at different subsites show different
patterns of lymphatic spread
\autocite{lindberg_distribution_1972,woolgar_histological_1999}. So far,
this could be handled by training different models for broader
categories of tumor locations, e.g.~oropharynx and oral cavity tumors.
However, this approach does not describe differences in lymphatic spread
between different subsites within the oropharynx and oral cavity. To
address this issue, we present an approach using mixtures of HMMs. The
intuition is that the lymphatic spread of a tumor that lies anatomically
at the boarder of oropharynx and oral cavity (e.g.~tumors in the palate)
may be described by a mixture of different models.

@@

In the above equation, the term
\(P \left( \mathbf{Y} \mid \mathbf{X} \right)\) is given by the
sensitivity and specificity of the diagnostic procedure. The term
\(P \left( \mathbf{X} \right)\) represents the prior probability of
involvement, which depends on the probability of the tumor to spread
through the lymphatic system. The main task of the HMM is to model
\(P \left( \mathbf{X} \right)\) and the main contribution of this paper
is to incorporate the primary tumor subsite into the model of
\(P \left( \mathbf{X} \right)\).

@@

A patient's state of lymph node involvement \(\mathbf{X}[t]\) evolves
over discrete time steps \(t\). Let us enumerate all \(2^V\) possible
states, representing all combinations of LNLs. In this paper, we
consider ipsilateral LNLs I, II, III and IV, which amounts to 16
possible states. The HMM is specified by a \emph{transition matrix}:

\begin{equation}\phantomsection\label{eq-transition-matrix}{
\mathbf{A} = \begin{pmatrix} A_{ij} \end{pmatrix} = P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right)
}\end{equation}

whose elements \(A_{ij}\) contain the conditional probabilities that a
state \(\mathbf{X}[t]=\boldsymbol{\xi}_i\) transitions to
\(\mathbf{X}[t+1]=\boldsymbol{\xi}_j\) over one time step. The
transition matrix is specified and parameterised via the graphical model
shown in figure~\ref{fig-schematic-and-graph}. The red arcs in the graph
of figure~\ref{fig-schematic-and-graph} (right panel) are associated
with the probability that the primary tumor spreads directly to a LNL
(parameters \(b_v\)). The blue arcs describe the spread from an upstream
LNL -- given it is already metastatic -- to a downstream level
(parameters \(t_{v \rightarrow v+1}\)).

Now, let \(\boldsymbol{\pi}\) be the \emph{starting distribution}

\begin{equation}\phantomsection\label{eq-starting-distribution}{
\boldsymbol{\pi} = \begin{pmatrix} \pi_i \end{pmatrix} = P \left( \mathbf{X}[0] = \boldsymbol{\xi}_i \right)
}\end{equation}

denoting the probability to start in state \(\boldsymbol{\xi}_i\) at
time step 0. Assuming that every patient started with all LNLs being
healthy, we set \(\pi_i\) to zero for all states except the completly
healthy state
\(\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}\), which
has probability one.

Using the quantities introduced so far, the probability
\(P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)\) to be in state
\(\boldsymbol{\theta}_i\) in time step \(t\) can now be conveniently
expressed as a matrix product:

\begin{equation}\phantomsection\label{eq-evolution}{
P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right) = \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
}\end{equation}

This evolution implicitly marginalizes over all possible paths to arrive
at state \(\boldsymbol{\xi}_i\) after \(t\) time-steps. Additionally, we
must marginalize over the unknown time of diagnosis using a time-prior
\(P_T(t)\). This finally defines the probability distribution over all
states of lymph node involvement used in equation~\ref{eq-bayes-law}.

@@

where \(\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}\) denotes
the set of all model parameters (7 in our case). Fortunately, the exact
length and shape of this distribution has little impact as previously
shown \autocite{ludwig_hidden_2021}. We set \(t_\text{max}=\) 10 and
\(P_\text{early}(t)\) to a binomial distribution with parameter 0.3.
Further details on the HMM can be found in
\textcite{ludwig_hidden_2021}.

@@

\centering{

\includegraphics{manuscript_files/static/schematic_and_graph.pdf}

}

\caption{\label{fig-schematic-and-graph}On the left: Anatomical sketch
of the tumor subsites and correspsonding ICD-10 codes considered in this
work. The subsite ``other parts of mouth'' (C06) was not drawn. On the
right: Parametrized graphical model of the lymphatic network considering
four LNLs. Blue nodes represent the hidden states of LNLs \(X_v\), while
the red one is the tumor. Arcs represent possible routes of metastatic
spread, associated with a probability.}

@@

Let us assume that we have a dataset \(\mathbf{D}\) that is specified
via the number of patients \(N_{is}\) that were diagnosed in LNL
involvement state \(i\) and had a primary tutor in subsite \(s\). Let us
further assume that we want to describe this dataset using a mixture of
\(M\) HMMs, each with a different set of model parameters
\(\boldsymbol{\theta_m}\). As the generative model of the data, we
assume that a patient with subsite \(s\) is generated with probability
\(\pi_{sm}\) from model \(m\). The likelihood of the dataset can then be
written as

\begin{equation}\phantomsection\label{eq-mixture-distribution}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_{sm} P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m \right) \right]^{N_{is}}
}\end{equation}

We now have two types of parameters, the probabilities of tumor spread
for the different models, \(\boldsymbol{\theta_m}\), and the mixing
coefficients \(\pi_{sm}\). Assuming a uniform prior in the interval
\([0,1]\) for all parameters, the posterior distribution over the
parameters
\(P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)\)
is given by the likelihood in equation~\ref{eq-mixture-distribution}
except for a normalisation constant. In this work, we use Markov chain
Monte Carlo sampling (MCMC) via the \texttt{emcee} Python package
\autocite{foreman-mackey_emcee_2013} to sample model parameters from the
posterior distribution. However,
\(P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)\)
itself is a multi-model distribution because one can permute the
different models. To address this problem, we revert to an
\emph{expectation-maximization (EM)} algorithm where we iterate two
steps until convergence of the mixing coefficients. In the E-step, we
sample model parameters \(\boldsymbol{\theta_m}\) using MCMC for given
mixing coefficients \(\pi_{sm}\). In the M-step, we maximize the
likelihood with respect to the mixing coefficients for given samples of
\(\boldsymbol{\theta_m}\)

@@

\centering{

\includegraphics{manuscript_files/figures/prevalence_by_subsite.pdf}

}

@@

In figure~\ref{fig-prevalence-by-subsite}, we plot the prevalence of
involvement in the four ipsilateral LNLs I, II, III, and IV stratified
by the primary tumor's subsite. The figure illustrates the variations in
LNL involvement between subsites within the oral cavity and oropharynx
categories. The number of patients for each subsite is indicated in
figure~\ref{fig-cluster-assignments}.

@@

We demonstrate the methodology for a mixture model with \(M=2\)
components, considering the ipsilateral involvement of LNLs I, II, III,
and IV and the primary tumor subsites shown in
figure~\ref{fig-prevalence-by-subsite}. In
figure~\ref{fig-cluster-assignments} we show the resulting mixture
coefficients \(\pi_{sm}\). The interpretation of this result is as
follows: tumors of the base of tongue (C01) are fully described by
component A, and tumors of the gum (C03) are fully described by
component B. These two subsites are the most distinct regarding the
involvement of LNLs I and II, and the result is thus intuitive.
Component A may be interpreted as a model for oropharynx-like tumor
spread, and component B as a model for oral cavity-like tumor spread.
All other subsites are described as mixtures. tumors in the tonsil (C09)
have LNL involvement similar to base of tongue tumors and are mostly
assigned to component A. Instead, tumors of the palate (C05) are to
similar degree assigned to components A and B, which is consistent with
the anatomical location and the observation that the LNL involvement is
in between oropharynx and oral cavity-type patterns
(figure~\ref{fig-prevalence-by-subsite}).

@@

\centering{

\includegraphics{manuscript_files/figures/cluster_assignments.pdf}

}

\caption{\label{fig-cluster-assignments}Assignment of each subsite to
each of the two components. The further left a subsite, the more it is
assigned to component A, the further right, the more to component B. The
size of the marker (area) corresponds to the number of patients in the
subsite.}

@@

The figure~\ref{fig-prevalence-comparison} illustrates the model's
predictions of the overall prevalence of lymph node involvement in LNLs
I, II and III (filled histograms) for selected subsites, obtained by
summing the probabilities of states where the respecive level is
involved. The mixture model is compared to two independent HMM models
trained for oral cavity and oropharynx (by pooling the respective
subsites). The mixture model and the independent oropharynx model
perform similarly for tonsil tumors (C09), which is the largest patient
group, dominating the independent oropharynx model and the component B
in the mixture model. However, the mixture model better predicts the
higher prevalence in level II for base of tongue and palate tumors.

@@

\centering{

\includegraphics{manuscript_files/figures/prevalence_comparison.pdf}

}

>>>>

\author[%
1,2%
]{Roman Ludwig}
\author[%
1,2%
]{Julian Brönnimann}
\author[%
1,2%
]{Yoel Perez Haas}
\author[%
1,2%
]{Esmée Lauren Looman}
\author[%
2%
]{Panagiotis Balermpas}
\author[%
11%
]{Sergi Benavente}
\author[%
3,4,7%
]{Adrian Schubert}
\author[%
8%
]{Dorothea Barbatei}
\author[%
8%
]{Laurence Bauwens}
\author[%
2%
]{Jean-Marc Hoffmann}
%\author[%
%4,5%
%]{Sandrine Werlen}
\author[%
3%
]{Olgun Elicin}
\author[%
6,10%
]{Matthias Dettmer}
%\author[%
%9%
%]{Philippe Zrounba}
\author[%
2%
]{Bertrand Poumayou}
\author[%
4,5%
]{Roland Giger}
\author[%
8%
]{Vincent Grégoire}
\author[%
1,2%
]{Jan Unkelbach}

@@

Head and neck squamous cell carcinomas (HNSCC) frequently spread through
the lymphatic system
\autocite{lindberg_distribution_1972,woolgar_histological_1999}. Current
diagnostic imaging modalities are unable to detect microscopic lymph
node metastases \autocite{snyder_petct_2021,strohl_petct_2021}. 
To avoid nodal recurrences 
%\autocite{ho_decision_2014}
, large volumes in the neck, which are at risk of harbouring occult disease are irradiated electively. Guidelines about which lymph node levels (LNLs) to irradiate
\autocite{biau_selection_2019} are currently not based on a patient's
individual risk, but on the overall prevalence of nodal disease as
reported in the literature
\autocite{lindberg_distribution_1972,woolgar_histological_1999}.
To personalize the prediction of the risk for occult disease, given a
patient's individual diagnosis, we published

@@

Such a personalized risk prediction may allow clinicians to safely
reduce the elective clinical target volume (CTV-N) and thus reduce
side-effects that degrade the patient's quality of life without compromising treatment efficacy
\autocite{batth_practical_2014}.

Here, we extend the previous work by incorporating the primary tumor
location (specified as ICD-10 code) into the model of lymphatic tumor
progression, focusing on tumors in the oropharynx and the oral cavity.
HNSCC patients with primary tumors at different subsites show different
patterns of lymphatic spread
\autocite{lindberg_distribution_1972,woolgar_histological_1999}. So far,
this could be handled by training different models for broader
categories of tumor locations, e.g.~oropharynx and oral cavity tumours.
However, this approach does not describe differences in lymphatic spread
between different subsites within the oropharynx and oral cavity. To
address this issue, we present an approach using mixtures of HMMs. The
intuition is that the lymphatic spread of a tumor that lies anatomically
at the boarder of oropharynx and oral cavity (e.g.~tumours in the
palate) may be described by a mixture of different models.

@@

In the above equation, the term
\(P \left( \mathbf{Y} \mid \mathbf{X} \right)\) is given by the sensitivity and specificity of the diagnostic procedure. The term \(P \left( \mathbf{X} \right)\) represents the prior probability of involvement, which depends on the probability of the tumor to spread through the lymphatic system. The main task of the HMM is to model
\(P \left( \mathbf{X} \right)\) and the main contribution of this paper is to incorporate the primary tumor subsite into the model for
\(P \left( \mathbf{X} \right)\).

@@

A patient's state of lymph node involvement \(\mathbf{X}[t]\) evolves
over discrete time steps \(t\). Let us enumerate all \(2^V\) possible
states, representing all combinations of LNLs. In this paper, we
consider ipsilateral LNLs I, II, III and IV, which amounts to 16
possible states. The HMM is specified by a \emph{transition matrix}
\begin{equation}\phantomsection\label{eq-transition-matrix}{
\mathbf{A} = \begin{pmatrix} A_{ij} \end{pmatrix} = P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right)
}\end{equation}

whose elements \(A_{ij}\) contains the conditional probabilities that a state
\(\mathbf{X}[t]=\boldsymbol{\xi}_i\) transitions to
\(\mathbf{X}[t+1]=\boldsymbol{\xi}_j\) over one time step. The
transition matrix is specified and parameterised via the graphical model
shown in figure~\ref{fig-schematic-and-graph}. The red arcs in the graph
of figure~\ref{fig-schematic-and-graph}(right panel) are associated with the probability that the primary tumor spreads directly to a LNL (parameters \(b_v\)).
The blue arcs describe the spread from an upstream LNL -- given it is
already metastatic -- to a downstream level (parameters
\(t_{v \rightarrow v+1}\)). \\

Now, let \(\boldsymbol{\pi}\) be the \emph{starting distribution}
\begin{equation}\phantomsection\label{eq-starting-distribution}{
\boldsymbol{\pi} = \begin{pmatrix} \pi_i \end{pmatrix} = P \left( \mathbf{X}[0] = \boldsymbol{\xi}_i \right)
}\end{equation}

denoting the probability to start in state \(\boldsymbol{\xi}_i\) at
time step 0. Assuming that every patient started with all
LNLs being healthy, we set \(\pi_i\) to zero for all states except the
completly healthy state
\(\boldsymbol{\xi} = \begin{pmatrix} 0, 0, 0, 0 \end{pmatrix}\), which
has probability one.
Using the quantities introduced so far, the probability
\(P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right)\) to be in state
\(\boldsymbol{\xi}_i\) in time step \(t\) can now be conveniently expressed as a matrix product:
\begin{equation}\phantomsection\label{eq-evolution}{
P \left( \mathbf{X}[t]=\boldsymbol{\xi}_i \right) = \left( \boldsymbol{\pi} \cdot \mathbf{A}^t \right)_i
}\end{equation}

This evolution implicitly marginalizes over all possible paths to arrive
at state \(\boldsymbol{\xi}_i\) after \(t\) time-steps. Additionally, we
must marginalize over the unknown time of diagnosis using a time-prior
\(P_T(t)\). This finally defines the probability distribution over all
states of lymph node involvement used in equation~(\ref{eq-bayes-law}).

@@

where \(\boldsymbol{\theta}=\{ b_v, t_{v \rightarrow v+1} \}\) denotes the
set of all model parameters (7 in our case). Fortunately, the exact
length and shape of this distribution has little impact as
previously shown \autocite{ludwig_hidden_2021}. We set \(t_\text{max}=\) 10 and \(P_\text{early}(t)\)
to a binomial distribution with parameter 0.3. Further details on the
HMM can be found in \autocite{ludwig_hidden_2021}.

@@

\centering{
\includegraphics{manuscript_files/static/schematic_and_graph.pdf}
}

\caption{\label{fig-schematic-and-graph}On the left: Anatomical sketch of the tumor subsites and correspsonding ICD-10 codes considered in this work. The subsite ``other parts of mouth'' (C06) was
not drawn. On the right: Parametrized graphical model of the
lymphatic network considering four LNLs. Blue nodes represent the hidden states of LNLs \(X_v\), while the red one is the tumor. Arcs represent possible routes of metastatic spread, associated with a probability.}
\end{figure}%

@@

Let us assume that we have a dataset \(\mathbf{D}\) that is specified
via the number of patients \(N_{is}\) that were diagnosed in LNL
involvement state \(i\) and had a primary tumor in subsite \(s\). Let us
further assume that we want to describe this dataset using a mixture of
\(M\) HMMs, each with a different set of model parameters
\(\boldsymbol{\theta_m}\). As the generative model of the data, we
assume that a patient with subsite \(s\) is generated with probability
\(\pi_{sm}\) from model \(m\). The likelihood of the dataset can then be
written as
\begin{equation}\phantomsection\label{eq-mixture-distribution}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_s \prod_i \left[ \sum_{m=1}^M \pi_{sm} P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m \right) \right]^{N_{is}}
}\end{equation}

We now have two types of parameters, the probabilities of tumor spread
for the different models, \(\boldsymbol{\theta_m}\), and the mixing
coefficients \(\pi_{sm}\). Assuming a uniform prior in the interval
\([0,1]\) for all parameters, the posterior distribution over the
parameters
\(P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)\)
is given by the likelihood in equation~(\ref{eq-mixture-distribution}) except
for a normalisation constant. In this work, we use Markov chain Monte
Carlo sampling (MCMC) via the \texttt{emcee} Python package
\autocite{foreman-mackey_emcee_2013} to sample model parameters from the
posterior distribution. However,
\(P \left( \boldsymbol{\theta}, \boldsymbol{\pi} \mid \mathbf{D} \right)\)
itself is a multi-model distribution because one can permute the
different models. To address this problem, we revert to an
\emph{expectation-maximization (EM)} algorithm where we iterate two steps until convergence of the mixing coefficients. In the E-step, we sample model parameters \(\boldsymbol{\theta_m}\) using MCMC for given mixing coefficients \(\pi_{sm}\). In the M-step, we maximize the likelihood with respect to the mixing coefficients for given samples of \(\boldsymbol{\theta_m}\).

@@

\centering{
\includegraphics{manuscript_files/figures/prevalence_by_subsite.pdf}
}

@@

In figure~\ref{fig-prevalence-by-subsite}, we plot the
prevalence of involvement in the four ipsilateral LNLs I, II, III, and
IV stratified by primary tumor subsite. 
%We included patients with tumors in the oral cavity, gum (C03), floor of mouth (C04), other/unspecified parts of the mouth (C06), other/unspecified parts of tongue (C02), palate (C05), and tumors in the oropharynx (C10), tonsil (C09), base of tongue (C01),  
The figure illustrates the variations in LNL involvement between subsites within
the oral cavity and oropharynx categories. The number of patients for each subsite is indicated in figure \ref{fig-cluster-assignments}.

@@

We demonstrate the methodology for a mixture model with \(M=2\)
components, considering the ipsilateral involvement of LNLs I, II, III,
and IV and the primary tumor subsites shown in
figure~\ref{fig-prevalence-by-subsite}. In
figure~\ref{fig-cluster-assignments} we show the resulting mixture
coefficients \(\pi_{sm}\). The interpretation of this result is as
follows: Tumors of the base of tongue (C01) are fully described by
component B, and tumors of the gum (C03) are fully described by
component A. These two subsites are the most distinct regarding the
involvement of LNLs I and II, and the result is thus intuitive.
Component B may be interpreted as a model for oropharynx-like tumor
spread, and component A as a model for oral cavity-like tumor spread.
All other subsites are described as mixtures. Tumors in the tonsil (C09)
have LNL involvement similar to base of tongue tumors and are mostly
assigned to component B. Instead, tumors of the palate (C05) are to
similar degree assigned to components A and B, which is consistent with
the anatomical location and the observation that the LNL involvement is
in between oropharynx and oral cavity-type patterns (figure~\ref{fig-prevalence-by-subsite}).

@@

\centering{
\includegraphics{manuscript_files/figures/cluster_assignments.pdf}
}

\caption{\label{fig-cluster-assignments}Assignment of each subsite to
each of the two components. The further left a subsite, the more it is
assigned to component A, the further right, the more to component B. The
size of the marker (area) corresponds to the number of patients in the
subsite indicated in brackets. The approximate locations of subsites are skeched in Figure \ref{fig-schematic-and-graph}.}

@@

Figure \ref{fig-prevalence-comparison} illustrates the model's predictions of the overall prevalence of lymph node involvement in LNLs I, II and III (filled histograms) for selected subsites, obtained by summing the probabilities of states where the respecive level is involved. The mixture model is compared to two independent HMM models trained for oral cavity and oropharynx (by pooling the respective subsites). The mixture model and the independent oropharynx model perform similarly for tonsil tumors (C09), which is the largest patient group, dominating the independent oropharynx model and the component B in the mixture model. However, the mixture model better predicts the higher prevalence in level II for base of tongue and palate tumors.

@@

\centering{
\includegraphics{manuscript_files/figures/prevalence_comparison.pdf}
}
